Accessing biological data with semantic web technologies

### 2013-12-30 article.tex
All SNP data is now converted to RDF

\begin{verbatim}
install.packages(c("rrdf", "rrdflibs"))
library(rrdf)
\end{verbatim}

@@

\begin{verbatim}
ontStore = new.rdf()
store = new.rdf(ontology=FALSE)
\end{verbatim}

@@

\begin{verbatim}
add.triple(store,
  subject="http://example.org/Subject",
  predicate="http://example.org/Predicate",
  object="http://example.org/Object"
)
add.data.triple(store,
  subject="http://example.org/Subject",
  predicate="http://example.org/Predicate",
  data="Literal value"
)
\end{verbatim}

@@

\begin{verbatim}
store = load.rdf("file.n3", format="N3")
\end{verbatim}

@@

\begin{verbatim}
library(biomaRt)
mart = biomaNoy2009Rt::useMart(biomart="snp", dataset="hsapiens_snp")
brca1 = c("rs16940","rs16941", "rs16942", "rs799916", "rs799917")
attribs = c(
  "refsnp_id", "chr_name", "chrom_start",
  "validated", "pmid_20137", "allele"
)
data = biomaRt::getBM(
  attributes=attribs, filters=c("snp_filter"), values=brca1, mart=mart
)
\end{verbatim}

@@

\begin{table}
\caption{A list of ontology terms for the concepts in the BRCA1 SNP example. Ontology acronyms and the prefixes
used are explained in the text.}
\begin{center}
\begin{tabular}{l|l|l|l}
 \textbf{concept} & \textbf{ontology} & \textbf{term} & \textbf{URI} \\
\hline
 SNP & SIO & snp & sio:SIO\_010027 \\
 gene & SO & gene & obo:SO\_0000704 \\
 chromosome & GO & chromosome & obo:GO\_0005694 \\
 article &  BIBO & article & bibo:Article \\
 allele & SO & allele & obo:SO\_0001023 \\
 SNP identifier & TMO & snp identifier & tmo:TMO\_0161 \\
 PubMed identifier & CHEMINF & PubMed identifier & sio:CHEMINF\_000302\\
 chromosome number & TMO & chromosome nuber & tmo:TMO\_0157 \\
 chromosomal sequence position & TMO & chromosomal sequence position & tmo:TMO\_0122 \\
\end{tabular}CiTO
\end{center}
\end{table}

Taking advantage of existing ontologies. The choice is somewhat arbitrary and no ontological
analysis has been done, e.g. on specified domain and ranges of predicates and implied
properties of classes. The following ontologies are used:
SIO is the Semanticscience Integrated Ontology; TMO is the Translational Medicine Ontology;
SO is the Sequence Types and Features Ontology; GO is the Gene Ontology;
BIBO is the Bibliographic Ontology; CiTO is the Citation Typing Ontology;
... Prefixes used in the term Uniform Resource Identifiers (URIs) include
\textit{sio} for \url{http://semanticscience.org/resource/},
\textit{tmo} for \url{http://www.w3.org/2001/sw/hcls/ns/transmed/},
\textit{obo} for \url{http://purl.obolibrary.org/obo/},
\textit{bibo} for \url{http://purl.org/ontology/bibo/},
\textit{cito} for \url{} ...

@@

\begin{verbatim}
snpStore = new.rdf(ontology=FALSE)
rdfTypePred = "http://www.w3.org/1999/02/22-rdf-syntax-ns#type"
snpClass = "http://semanticscience.org/resource/SIO_010027"
createEntry <- function(row) {
  snpID = row[1]
  snpSubject = paste("http://example.org/snp/", snpID, sep="")
  add.triple(snpStore,
    subject=snpSubject, predicate=rdfTypePred, object=snpClass
  )
}
\end{verbatim}

We can then just iterate over the actual data with:

\begin{verbatim}
apply(data, MARGIN=1, FUN=createEntry(row) )
\end{verbatim}

@@

\begin{verbatim}
sparql.rdf(store,
  paste(
    "SELECT DISTINCT ?type WHERE {",
    " [] a ?type ",
    "}"
  )
) 
\end{verbatim}

@@

\begin{verbatim}
results = sparql.remote(
  "http://rdf.farmbio.uu.se/chembl/sparql",
  paste(
    "SELECT DISTINCT ?predicate ?object WHERE {",
    " ?assay <http://www.w3.org/2000/01/rdf-schema#label> \"CHEMBL615603\" ;",
    "        ?predicate ?object .",
    "}"
  )
) 
\end{verbatim}

@@

\begin{verbatim}
results[,"predicate"]
\end{verbatim}

>>>>

\begin{footnotesize}
\begin{verbatim}
install.packages(c("rrdf", "rrdflibs"))
library(rrdf)
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
ontStore = new.rdf()
store = new.rdf(ontology=FALSE)
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
add.triple(store,
  subject="http://example.org/Subject",
  predicate="http://example.org/Predicate",
  object="http://example.org/Object"
)
add.data.triple(store,
  subject="http://example.org/Subject",
  predicate="http://example.org/Predicate",
  data="Literal value"
)
\end{verbatim}
\end{footnotesize}

We can also take advantage from the fact that R assigns values to method parameters based on
the order in which they are given. Therefore, we can simplify the above code to:

\begin{footnotesize}
\begin{verbatim}
add.triple(store,
  "http://example.org/Subject",
  "http://example.org/Predicate",
  "http://example.org/Object"
)
add.data.triple(store,
  "http://example.org/Subject",
  "http://example.org/Predicate",
  "Literal value"
)
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
store = load.rdf("file.n3", format="N3")
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
library(biomaRt)
mart = biomaNoy2009Rt::useMart(biomart="snp", dataset="hsapiens_snp")
brca1 = c("rs16940","rs16941", "rs16942", "rs799916", "rs799917")
attribs = c(
  "refsnp_id", "chr_name", "chrom_start",
  "validated", "pmid_20137", "allele"
)
data = biomaRt::getBM(
  attributes=attribs, filters=c("snp_filter"), values=brca1, mart=mart
)
\end{verbatim}
\end{footnotesize}

@@

\begin{table}
\caption{A list of ontology terms for the concepts in the BRCA1 SNP example. Ontology acronyms and the prefixes
used are explained in the text.}
\begin{center}
\begin{footnotesize}
\begin{tabular}{l|l|l|l}
 \textbf{concept} & \textbf{ontology} & \textbf{term} & \textbf{URI} \\
\hline
 SNP & SIO & snp & sio:SIO\_010027 \\
 chromosome & GO & chromosome & obo:GO\_0005694 \\
 article &  BIBO & article & bibo:Article \\
 allele & SO & allele & obo:SO\_0001023 \\
 SNP identifier & TMO & snp identifier & tmo:TMO\_0161 \\
 PubMed identifier & CHEMINF & PubMed identifier & sio:CHEMINF\_000302\\
 chromosome number & TMO & chromosome nuber & tmo:TMO\_0157 \\
 chromosomal sequence position & TMO & chromosomal sequence position & tmo:TMO\_0122 \\
\end{tabular}
\end{footnotesize}
\end{center}
\end{table}

Taking advantage of existing ontologies. The choice is somewhat arbitrary and no ontological
analysis has been done, e.g. on specified domain and ranges of predicates and implied
properties of classes. The following ontologies are used:
SIO is the Semanticscience Integrated Ontology;
TMO is the Translational Medicine Ontology;
%SO is the Sequence Types and Features Ontology;
GO is the Gene Ontology;
BIBO is the Bibliographic Ontology;
CiTO is the Citation Typing Ontology;
... Prefixes used in the term Uniform Resource Identifiers (URIs) include
\textit{sio} for \url{http://semanticscience.org/resource/},
\textit{tmo} for \url{http://www.w3.org/2001/sw/hcls/ns/transmed/},
\textit{obo} for \url{http://purl.obolibrary.org/obo/},
\textit{bibo} for \url{http://purl.org/ontology/bibo/},
\textit{cito} for \url{http://purl.org/spar/cito/} ...

We can define these prefixes in the store with the following code,
the first prefix in the full syntax and the other namespaces
leaving out the parameters names:

\begin{footnotesize}
\begin{verbatim}
snpStore = new.rdf(ontology=FALSE)
add.prefix(snpStore,
  prefix="sio", namespace="http://semanticscience.org/resource/"
)
add.prefix(snpStore, "tmo", "http://www.w3.org/2001/sw/hcls/ns/transmed/")
add.prefix(snpStore, "obo", "http://purl.obolibrary.org/obo/")
add.prefix(snpStore, "bibo", "http://purl.org/ontology/bibo/")
add.prefix(snpStore, "cito", "http://purl.org/spar/cito/")
add.prefix(snpStore, "snp", "http://example.org/snp/")
add.prefix(snpStore, "art", "http://example.org/article/")
add.prefix(snpStore, "loc", "http://example.org/location/")
add.prefix(snpStore, "all", "http://example.org/allele/")
add.prefix(snpStore, "ex", "http://example.org/onto/")
add.prefix(snpStore, "pubmedid", "http://example.org/pubmed/")
add.prefix(snpStore, "snpid", "http://example.org/snpid/")
\end{verbatim}
\end{footnotesize}

Furthermore, we can define a number of classes and predicate, for reduced code:

\begin{footnotesize}
\begin{verbatim}
snpClass = "http://semanticscience.org/resource/SIO_010027"
chromosomeClass = "http://purl.obolibrary.org/obo/GO_0005694"
pubmedIdClass = "http://semanticscience.org/resource/CHEMINF_000302"
snpIdClass = "http://www.w3.org/2001/sw/hcls/ns/transmed/TMO_0161"
articleClass = "http://purl.org/ontology/bibo/Article"
alleleClass = "http://purl.obolibrary.org/obo/SO_0001023"

rdfTypePred = "http://www.w3.org/1999/02/22-rdf-syntax-ns#type"
isDescribedBy = "http://purl.org/spar/cito/isDescribedBy"
onChromosome = "http://example.org/onto/onChromosome"
hasAttribute = "http://semanticscience.org/resource/CHEMINF_000200"
hasStart = "http://example.org/onto/hasStart"
hasValue = "http://semanticscience.org/resource/SIO_000300"
hasValidation = "http://example.org/onto/hasValidation"
hasAllele = "http://example.org/onto/hasAllele"
hasLocation = "http://example.org/onto/hasLocation"
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
createEntry <- function(row) {
  snpID = row[1]
  chr = row[2]
  chrStart = row[3]
  validation = row[4]
  pubmedID = row[5]
  allele = row[6]

  snpSubject = paste("http://example.org/snp/", snpID, sep="")
  pubmedObject = paste("http://example.org/article/a", pubmedID, sep="")
  alleleObject = paste("http://example.org/allele/", snpID, sep="")
  chrObject = paste("http://example.org/location/", snpID, sep="")

  snpIDObject = paste("http://example.org/snpid/", snpID, sep="")
  pubmedIDObject = paste("http://example.org/pubmed/a", pubmedID, sep="")

  add.triple(snpStore, snpSubject, rdfTypePred, snpClass)
  add.triple(snpStore, snpSubject, hasAttribute, snpIDObject)
  add.triple(snpStore, snpIDObject, rdfTypePred, snpIdClass)
  add.data.triple(snpStore, snpIDObject, hasValue, snpID)
  # validation information
  sapply(unlist(strsplit(validation, split=",")),
    function(validationItem) {
      if (!is.null(validationItem)) {
        add.data.triple(snpStore,
          snpSubject, hasValidation, validationItem
        )
      }
    }
  )
  # allele information
  add.triple(snpStore, snpSubject, hasAllele, alleleObject)
  add.triple(snpStore, alleleObject, rdfTypePred, alleleClass)
  add.data.triple(snpStore, alleleObject, hasValue, allele)
  # chromosome location information
  add.triple(snpStore, snpSubject, hasLocation, chrObject)
  add.triple(snpStore, chrObject, rdfTypePred, chromosomeClass)
  add.data.triple(snpStore, chrObject, onChromosome, chr)
  add.data.triple(snpStore, chrObject, hasStart, chrStart)
  # pubmed information
  add.triple(snpStore, snpSubject, isDescribedBy, pubmedObject)
  add.triple(snpStore, pubmedObject, rdfTypePred, articleClass)
  add.triple(snpStore, pubmedObject, hasAttribute, pubmedIDObject)
  add.triple(snpStore, pubmedIDObject, rdfTypePred, pubmedIdClass)
  add.data.triple(snpStore, pubmedIDObject, hasValue, pubmedID)
}
\end{verbatim}
\end{footnotesize}

We can then just iterate over the actual data with and safe it to a file:

\begin{footnotesize}
\begin{verbatim}
apply(data, MARGIN=1, FUN=createEntry)
save.rdf(snpStore, filename="test.n3", format="N3")
\end{verbatim}
\end{footnotesize}

Or output is as a Notation3 string with:

\begin{footnotesize}
\begin{verbatim}
cat(asString.rdf(snpStore))
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
sparql.rdf(store,
  paste(
    "SELECT DISTINCT ?type WHERE {",
    " [] a ?type ",
    "}"
  )
) 
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
results = sparql.remote(
  "http://rdf.farmbio.uu.se/chembl/sparql",
  paste(
    "SELECT DISTINCT ?predicate ?object WHERE {",
    " ?assay <http://www.w3.org/2000/01/rdf-schema#label> \"CHEMBL615603\" ;",
    "        ?predicate ?object .",
    "}"
  )
) 
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
results[,"predicate"]
\end{verbatim}
\end{footnotesize}
